\textbf{SPECFEM3D}\\ \textbf{Cartesian}\\ \textbf{User Manual}

### 2022-11-09 doc/USER_MANUAL/04_creating_databases.tex
adds option (MOVIE_TYPE == 3) for shakemap to output geometric mean of horizontal PG-values
Path updates for parameters moved from constants.h to Par_file and vice versa, and other minor corrections.

\item [{\texttt{APPROXIMATE\_OCEAN\_LOAD}}] Set to \texttt{.true.} if the
effect of the oceans on seismic wave propagation should be incorporated
based upon the (rough) approximate treatment discussed in \citet{KoTr02b}.
This feature is inexpensive from a numerical perspective, both in
terms of memory requirements and CPU time. This approximation is accurate
at periods of roughly 20~s and longer. At shorter periods the effect
of water phases/reverberations is not taken into account, even when
the flag is on. If you want to model the effect of a fluid-solid model
at short periods, then set this flag to \texttt{.false.} and mesh
the fluid layer explicitly in your mesher, so that it is computed
accurately and without this approximation.
\item [{\texttt{TOPOGRAPHY}}] This feature is only effective if \texttt{APPROXIMATE\_OCEAN\_LOAD}
is set to \texttt{.true.}. Set to \texttt{.true.} if topography and
bathymetry should be read in based upon the topography file specified
in the main constants file \texttt{constants.h} found in subdirectory
\texttt{src/shared/} to evaluate elevations. If not set, elevations
will be read from the numerical mesh.
\item [{\texttt{ATTENUATION}}] Set to \texttt{.true.} if attenuation should
be incorporated. Turning this feature on increases the memory requirements
significantly (roughly by a factor of~1.5), and is numerically fairly
expensive. See \citet{KoTr99,KoTr02a} for a discussion on the implementation
of attenuation based upon standard linear solids. Please note that
the Vp- and Vs-velocities of your model are given for a reference
frequency. To change this reference frequency, you change the value
of \texttt{ATTENUATION\_f0\_REFERENCE} in the main constants file
\texttt{constants.h} found in subdirectory \texttt{src/shared/}.
The code uses a constant $Q$ quality factor, write(IMAIN,*) "but approximated based on a series of Zener standard linear solids (SLS).
The approximation is thus performed in a given frequency band determined based on that \texttt{ATTENUATION\_f0\_REFERENCE} reference frequency.
Note that Qmu is always equal to Qs, but Qkappa is in general not equal to Qp.
To convert one to the other see \texttt{doc/note\_on\_Qkappa\_versus\_Qp.pdf}
and in folder \texttt{utils/attenuation}, the tool
\texttt{conversion\_from\_Qkappa\_Qmu\_to\_Qp\_Qs\_from\_Dahlen\_Tromp\_959\_960.f90}.

@@

create a ShakeMap\textregistered{}, i.e., a peak ground velocity map

@@

When using this option, by default the code can locate the force source
anywhere between mesh points in order to honor its exact location; this
is more precise than using the closest GLL mesh point, but it is also a bit slower.
If needed, you can change that default behavior and force the code to use the closest
GLL mesh point instead by setting flag \texttt{USE\_BEST\_LOCATION} to \texttt{.false.} instead of \texttt{.true.} in file
\texttt{src/shared/constants.h.in} and running the \texttt{configure} script again and recompiling the code.
\item [{\texttt{USE\_RICKER\_TIME\_FUNCTION}}] Turn this flag on to use
a Ricker source time function, i.e., the second derivative of a Gaussian, instead of the source time functions
set by default to represent a (tilted) \texttt{FORCESOLUTION} force
point source or a \texttt{CMTSOLUTION} moment-tensor source.
Note that we use the standard definition of a Ricker, for a dominant frequency $f_0$:
$\mathrm{Ricker}(t) = (1 - 2 a t^2) e^{-a t^2}$, with $a = \pi^2 f_0^2$,
whose Fourier transform is thus:
$\frac{1}{2} \frac{\sqrt{\pi}\omega^2}{a^{3/2}}e^{-\frac{\omega^2}{4 a}}$
This gives the wavelet of Figure~\ref{fig:RickerWavelet}.
%%
\begin{figure}[htbp]
\centering
\includegraphics[width=3in]{figures/Ricker_wavelet.png}
\caption{We use the standard definition of a Ricker (i.e., second derivative of a Gaussian). Image taken from \url{http://subsurfwiki.org}.}
\label{fig:RickerWavelet}
\end{figure}
%%
Originally, if a \texttt{CMTSOLUTION} moment-tensor source is used, a (pseudo)
Heaviside step function with a very short half duration is defined
for elastic cases to represent the permanent slip on the fault while
in the acoustic case a Gaussian source time function with a similarly
short half duration is defined to physically describe actions within
the fluid. Otherwise, if a \texttt{FORCESOLUTION} force source is
used, a (pseudo) Dirac delta source time function is defined by default.
Any other source-time function may then be obtained by convolution.
\item [{\texttt{PRINT\_SOURCE\_TIME\_FUNCTION}}] Turn this flag on to print
information about the source time function in the file \texttt{OUTPUT\_FILES/plot\_source\_time\_function.txt}.
This feature is only relevant for the solver.
%
\end{description}

@@

When that option is on, of course the number of processor cores used to start
the code in the batch system must be a multiple of \texttt{NUMBER\_OF\_SIMULTANEOUS\_RUNS},
all the individual runs must use the same number of processor cores,
which as usual is NPROC in the Par\_file,
and thus the total number of processor cores to request from the batch system
should be \texttt{NUMBER\_OF\_SIMULTANEOUS\_RUNS}~*~\texttt{NPROC}.
All the runs to perform must be placed in directories called run0001, run0002, run0003 and so on
(with exactly four digits).

@@

\item [\texttt{ADIOS\_ENABLED}] Turn this flag on to enable ADIOS. If set to \texttt{.false.}, subsequent ADIOS
parameters will not be considered.
\item [\texttt{ADIOS\_FOR\_DATABASES}] Turn this flag on to use ADIOS for xmeshfem3D output and
xgenerate\_database input.
\item [\texttt{ADIOS\_FOR\_MESH}]  Turn this flag on to use ADIOS for generated databases.
\item [\texttt{ADIOS\_FOR\_FORWARD\_ARRAYS}] Turn this flag on to read and write forward arrays using ADIOS.
\item [\texttt{ADIOS\_FOR\_KERNELS}] Turn this flag on to produce ADIOS kernels that can later be visualized with the ADIOS version of combine\_vol\_data.
\end{description}
There are quite a few more parameters to change the default setup of your runs. Please check the comments in the \texttt{Par\_file} directly for further explanations.\\

@@

The parameter \texttt{DT} sets the length of each time step in seconds.
The value of this parameter is crucial for the stability of the spectral-element
simulation. Your time step \texttt{DT} will depend on the minimum
ratio between the distance $h$ of neighboring mesh points and the
wave speeds $v$ defined in your model. The condition for the time
step $\Delta t$ is:
\begin{lyxcode}
$\Delta t<C~\mathrm{min}_{\Omega}(~h/v~)$
\end{lyxcode}
where $C$ is the so-called Courant number and $\Omega$ denotes the
model volume. The distance $h$ depends on the mesh element size and
the number of GLL points \texttt{NGLL} specified in the main constants
file \texttt{constants.h} located in the \texttt{src/shared/} subdirectory.
The wave speed $v$ is determined based on your model's P- (or S-)
wave speed values.

>>>>

\item [{\texttt{APPROXIMATE\_OCEAN\_LOAD}}] Set to \texttt{.true.} if the
effect of the oceans on seismic wave propagation should be incorporated
based upon the (rough) approximate treatment discussed in \citet{KoTr02b}.
This feature is inexpensive from a numerical perspective, both in
terms of memory requirements and CPU time. This approximation is accurate
at periods of roughly 20~s and longer. At shorter periods the effect
of water phases/reverberations is not taken into account, even when
the flag is on. If you want to model the effect of a fluid-solid model
at short periods, then set this flag to \texttt{.false.} and mesh
the fluid layer explicitly in your mesher, so that it is computed
accurately and without this approximation.
\item [{\texttt{TOPOGRAPHY}}] This feature is only effective if \texttt{APPROXIMATE\_OCEAN\_LOAD}
is set to \texttt{.true.}. Set to \texttt{.true.} if topography and
bathymetry should be read in based upon the topography file specified
in the main constants file \texttt{setup/constants.h} to evaluate elevations. If not set, elevations
will be read from the numerical mesh.
\item [{\texttt{ATTENUATION}}] Set to \texttt{.true.} if attenuation should
be incorporated. Turning this feature on increases the memory requirements
significantly (roughly by a factor of~1.5), and is numerically fairly
expensive. See \citet{KoTr99,KoTr02a} for a discussion on the implementation
of attenuation based upon standard linear solids. Please note that
the Vp- and Vs-velocities of your model are given for a reference
frequency. To change this reference frequency, you change the value
of \texttt{ATTENUATION\_f0\_REFERENCE} defined (in Hz) in input file \texttt{DATA/Par\_file}.
The code uses a constant $Q$ quality factor, write(IMAIN,*) "but approximated based on a series of Zener standard linear solids (SLS).
The approximation is thus performed in a given frequency band determined based on that \texttt{ATTENUATION\_f0\_REFERENCE} reference frequency.
Note that Qmu is always equal to Qs, but Qkappa is in general not equal to Qp.
To convert one to the other see \texttt{doc/note\_on\_Qkappa\_versus\_Qp.pdf}
and in folder \texttt{utils/attenuation}, the tool
\texttt{conversion\_from\_Qkappa\_Qmu\_to\_Qp\_Qs\_from\_Dahlen\_Tromp\_959\_960.f90}.

@@

create a ShakeMap\textregistered{}, e.g., a peak ground velocity map

@@

When using this option, by default the code can locate the force source
anywhere between mesh points in order to honor its exact location; this
is more precise than using the closest GLL mesh point, but it is also a bit slower.
If needed, you can change that default behavior and force the code to use the closest
GLL mesh point instead by setting flag \texttt{USE\_BEST\_LOCATION} to \texttt{.false.} instead of \texttt{.true.} in file
\texttt{setup/constants.h} and recompiling the code.
\item [{\texttt{USE\_RICKER\_TIME\_FUNCTION}}] Turn this flag on to use
a Ricker source time function, i.e., the second derivative of a Gaussian, instead of the source time functions
set by default to represent a (tilted) \texttt{FORCESOLUTION} force
point source or a \texttt{CMTSOLUTION} moment-tensor source.
Note that we use the standard definition of a Ricker, for a dominant frequency $f_0$:
$\mathrm{Ricker}(t) = (1 - 2 a t^2) e^{-a t^2}$, with $a = \pi^2 f_0^2$,
whose Fourier transform is thus:
$\frac{1}{2} \frac{\sqrt{\pi}\omega^2}{a^{3/2}}e^{-\frac{\omega^2}{4 a}}$
This gives the wavelet of Figure~\ref{fig:RickerWavelet}.
%%
\begin{figure}[htbp]
\centering
\includegraphics[width=3in]{figures/Ricker_wavelet.png}
\caption{We use the standard definition of a Ricker (i.e., second derivative of a Gaussian). Image taken from \url{http://subsurfwiki.org}.}
\label{fig:RickerWavelet}
\end{figure}
%%
Originally, if a \texttt{CMTSOLUTION} moment-tensor source is used, a (pseudo)
Heaviside step function with a very short half duration is defined
for elastic cases to represent the permanent slip on the fault while
in the acoustic case a Gaussian source time function with a similarly
short half duration is defined to physically describe actions within
the fluid. Otherwise, if a \texttt{FORCESOLUTION} force source is
used, a (pseudo) Dirac delta source time function is defined by default.
Any other source-time function may then be obtained by convolution.
\item [{\texttt{PRINT\_SOURCE\_TIME\_FUNCTION}}] Turn this flag on to print
information about the source time function in the file \texttt{OUTPUT\_FILES/plot\_source\_time\_function.txt}.
This feature is only relevant for the solver.
%
\end{description}

@@

When that option is on, of course the number of processor cores used to start
the code in the batch system must be a multiple of \texttt{NUMBER\_OF\_SIMULTANEOUS\_RUNS},
all the individual runs must use the same number of processor cores,
which as usual is NPROC in the \texttt{DATA/Par\_file},
and thus the total number of processor cores to request from the batch system
should be \texttt{NUMBER\_OF\_SIMULTANEOUS\_RUNS}~*~\texttt{NPROC}.
All the runs to perform must be placed in directories called run0001, run0002, run0003 and so on
(with exactly four digits).

@@

\item [\texttt{ADIOS\_ENABLED}] Turn this flag on to enable ADIOS. If set to \texttt{.false.}, subsequent ADIOS
parameters will not be considered.
\item [\texttt{ADIOS\_FOR\_DATABASES}] Turn this flag on to use ADIOS for xmeshfem3D output and
xgenerate\_database input.
\item [\texttt{ADIOS\_FOR\_MESH}]  Turn this flag on to use ADIOS for generated databases.
\item [\texttt{ADIOS\_FOR\_FORWARD\_ARRAYS}] Turn this flag on to read and write forward arrays using ADIOS.
\item [\texttt{ADIOS\_FOR\_KERNELS}] Turn this flag on to produce ADIOS kernels that can later be visualized with the ADIOS version of combine\_vol\_data.
\end{description}
There are quite a few more parameters to change the default setup of your runs. Please check the comments in the \texttt{DATA/Par\_file} directly for further explanations.\\

@@

The parameter \texttt{DT} sets the length of each time step in seconds.
The value of this parameter is crucial for the stability of the spectral-element
simulation. Your time step \texttt{DT} will depend on the minimum
ratio between the distance $h$ of neighboring mesh points and the
wave speeds $v$ defined in your model. The condition for the time
step $\Delta t$ is:
\begin{lyxcode}
$\Delta t<C~\mathrm{min}_{\Omega}(~h/v~)$
\end{lyxcode}
where $C$ is the so-called Courant number and $\Omega$ denotes the
model volume. The distance $h$ depends on the mesh element size and
the number of GLL points \texttt{NGLL} specified in the main constants
file \texttt{setup/constants.h}. The wave speed $v$ is determined based on your model's P- (or S-)
wave speed values.
